The Power of Two Random Choices

### 2014-01-13 article/article.tex
move paragraph about heavily loaded case to introduction
applications: hashing and PRAM emulation on DMM

\section{Classification}
\label{sec:classification}
Several different variations of this problem exist, which may be classified by the following properties:
\begin{itemize}
\item In a \emph{sequential process}, the balls arrive one after another. A  ball is dealt with only after the previous ball has been placed. Contrary, in a \emph{parallel process}, balls arrive in sets and the balls in one set are dealt with simultaneously \cite{ABS98}. 
\item In a \emph{static problem}, the balls are placed into the bins and no balls are ever removed. In a \emph{dynamic problem}, balls can also be removed from the system. In a simple \emph{random deletions} model, for each newly inserted ball a ball choosen uniformly and independently at random from among the balls currently in the bins is removed from the system \cite{ABKU99} \cite{RMS01}. More elaborate shemes exists: An \emph{oblivious adversery} specifies the sequence of insertions and deletions of balls in advance, without any knowledge of the placement of the balls \cite{CFM+98}. Another approach was studied by Adler et. al \cite{ABS98} in the context of the parallel balls into bins problem: In each bin, the balls are stored in FIFO-order and at every time step the first ball in each of the bins is deleted. 
\item A \emph{finite process} is analyzed for a given time interval that is known beforehand, whereas for an \emph{infinite process}, a placement and a deletion process is defined over an infinite time span \cite{ABS98}.
\end{itemize}

\section{Applications}
\label{sec:applications}
\subsection{Hashing}
\subsection{On-line load balancing}
\subsection{Dynamic resource allocation}
\subsection{Shared memory emulation on distributed memory machines}

@@

This was first proven by Azar et al. \cite{ABKU99} for the case $m = n$. (However, the case $d = 2$ was implicitly proven by Karp et al. \cite{KLM92}). See also \cite{RMS01} for a simplified proof and an overview of proof techniques for the given problem.  Unfortunately, the analysis breaks down for the heavily loaded case, where $m = \omega\left(n\right)$. Berenbrink at. all \cite{BFZR08} were the first to proof that the same bound holds for the heavily loaded case\footnote{However, they say that an event $\mathcal E$ occurs with high probability if $\Pr\left[\mathcal E \right]  \geq 1- n^{-\alpha}$, for an arbitrarily chosen constant $\alpha \geq 1$.}. A simplified proof may be found in \cite{TW13}.

The result stated above implies a gap of $\Theta\left(\log \log n \right)$ for arbitrarily large $m$. This is remarkable in several ways:
\begin{itemize}
\item The apparently small change made for the \emph{multiple-choice balls into bins} strategy results in an exponential decrease of the gap between the fullest box and the average load, even for $d=2$. 
\item Each additional choice ($d > 2$) decreases the gap by only a constant factor \cite{RMS01}. 
\item For the \emph{heavily loaded case}, the resulting gap of the \emph{multiple-choice balls into bins} strategy does not depend on the number of balls $m$. In contrast, the bound for the \emph{one-choice} diverges with $m$.
\item The given bounds are \emph{tight}, meaning that no other strategy that places each ball into one of $d$ randomly selected bins achieves a gap that is asymptotically lower.
\end{itemize}

>>>>

\section{Classification}
\label{sec:classification}
Several different variations of this problem exist, which may be classified by the following properties:
\begin{itemize}
\item In a \emph{sequential process}, the balls arrive one after another. A  ball is dealt with only after the previous ball has been placed. Contrary, in a \emph{parallel process}, balls arrive in sets and the balls in one set are dealt with simultaneously \cite{ABS98}. 
\item In a \emph{static problem}, the balls are placed into the bins and no balls are ever removed. In a \emph{dynamic problem}, balls can also be removed from the system. In a simple \emph{random deletions} model, for each newly inserted ball a ball choosen uniformly and independently at random from among the balls currently in the bins is removed from the system \cite{ABKU99} \cite{MRS01}. More elaborate shemes exists: An \emph{oblivious adversery} specifies the sequence of insertions and deletions of balls in advance, without any knowledge of the placement of the balls \cite{CFM+98}. Another approach was studied by Adler et. al \cite{ABS98} in the context of the parallel balls into bins problem: In each bin, the balls are stored in FIFO-order and at every time step the first ball in each of the bins is deleted. 
\item A \emph{finite process} is analyzed for a given time interval that is known beforehand, whereas for an \emph{infinite process}, a placement and a deletion process is defined over an infinite time span \cite{ABS98}.
\end{itemize}

\section{Applications}
\label{sec:applications}
\subsection{Hashing}
A simple hash table implementation typically uses a single hash function to map elements to entries in a table. If more than one element is mapped to the same entry, all elements mapped to this table entry are stored in a \emph{chain}, usually implemented as a linked list. The maximum time to search for an element is then proportional to the length of the longest chain in the hash table.

It is easy to see how the balls-into-bins idea can be applied: Assuming a perfect hash function\footnote{i.e., each of the $n$ elements is mapped to one of the $n$ table entries uniformly and independently at random}, the elements are balls to be mapped to table entries or bins, implying a worst case look-up time of $\Theta\left(\frac{\log n}{\log \log n}\right)$ w.h.p. Applying the two-choice paradigm by using two hash functions and mapping elements to the least full of two table entries implies a bound of $\Theta\left(\log \log n\right)$ w.h.p. This approach is easy to parallelize, it does not require re-hashing of data and requires only two hash functions \cite{MRS01} \cite{ABKU99}.

\subsection{On-line load balancing}

\subsection{Dynamic resource allocation}

\subsection{Shared memory emulation on distributed memory machines}
The \emph{Parallel Random Access Machine} model\footnote{as defined in the lecture or in~\cite{P03}}, or \emph{PRAM} for short, is a high level abstraction of a shared memory machine, in which memory access by any processor to any memory location is assumed to require constant time. It is especially useful for the study of parallel algorithms. However, a PRAM is much harder to realize in hardware than a \emph{Distributed Memory Machine} (DMM). Thus, ways to emulate a PRAM on a DMM have been studied extensivly. Such an emulation distributes the processors and memory cells of the PRAM to the processors and memory modules (typically one per processor) of the DMM with the goal to reduce the slowdown of the emulation, which is the time the DMM requires to simulate one step of the PRAM. Each memory modul can handle only one memory access at a time step. More accesses to the same module will result in memory contention. The balls-into-bins idea is useful for managing communication between the processors and the memory modules such that memory contention is minimal. For details, see Karp et. al \cite{KLM92} and the papers referenced there.

@@

This was first proven by Azar et al. \cite{ABKU99} for the case $m = n$. (However, the case $d = 2$ was implicitly proven by Karp et al. \cite{KLM92}). See also \cite{MRS01} for a simplified proof and an overview of proof techniques for the given problem.  Unfortunately, the analysis breaks down for the heavily loaded case, where $m = \omega\left(n\right)$. Berenbrink at. all \cite{BFZR08} were the first to proof that the same bound holds for the heavily loaded case\footnote{However, they say that an event $\mathcal E$ occurs with high probability if $\Pr\left[\mathcal E \right]  \geq 1- n^{-\alpha}$, for an arbitrarily chosen constant $\alpha \geq 1$.}. A simplified proof may be found in \cite{TW13}.

The result stated above implies a gap of $\Theta\left(\log \log n \right)$ for arbitrarily large $m$. This is remarkable in several ways:
\begin{itemize}
\item The apparently small change made for the \emph{multiple-choice balls into bins} strategy results in an exponential decrease of the gap between the fullest box and the average load, even for $d=2$. 
\item Each additional choice ($d > 2$) decreases the gap by only a constant factor \cite{MRS01}. 
\item For the \emph{heavily loaded case}, the resulting gap of the \emph{multiple-choice balls into bins} strategy does not depend on the number of balls $m$. In contrast, the bound for the \emph{one-choice} diverges with $m$.
\item The given bounds are \emph{tight}, meaning that no other strategy that places each ball into one of $d$ randomly selected bins achieves a gap that is asymptotically lower.
\end{itemize}
